\section{Data}
\label{sec:data}

Our dataset comprises 30-minute intraday observations beginning in January 2005. The primary forecasting target is realized volatility, measured as the sum of squared returns within each 30-minute bar ($RV_t = \texttt{sumret2}$).

\subsection{Exogenous Variables}
\label{sec:exog_vars}

In addition to the target series, we incorporate variables spanning several information categories:

\begin{itemize}
    \item \textbf{Higher-order return moments}: bipower variation, realized semi-variance, autocovariance, absolute returns, cubic returns, and quartic returns.
    \item \textbf{Liquidity and microstructure}: trading volume, turnover (total, buy-initiated, sell-initiated), effective spread, quoted spread, and number of trades.
    \item \textbf{Cross-sectional stock factors}: equal-weighted and value-weighted aggregates of stock-level return moments, bipower variation, semi-variance, and absolute returns.
    \item \textbf{S\&P 500 and sentiment}: SPY trading activity (turnover, buy/sell turnover) and StockTwits social media metrics (attention, sentiment score, sentiment count).
    \item \textbf{Implied volatility}: VIX, VVIX, and VIX3M.
    \item \textbf{Volatility demand}: Options order flow indicators (SPX open/close, all options open/close, open only).
\end{itemize}

\subsection{Data Cleaning}
\label{sec:cleaning}

% TODO: Expand with additional context
The raw data is gridded to a regular 30-minute frequency and filtered to exclude non-trading periods: Friday evenings after 20:00, all of Saturday, and Sunday mornings before 18:30. Overnight gaps in stock factors and volatility demand measures are forward-filled with neutral values. After cleaning and lag feature generation (which requires a burn-in period of 3{,}125 bars for the longest HAR lag), the final sample comprises approximately 218{,}000 observations.
