\section{Methodology}
\label{sec:methodology}

This section details the full modeling pipeline: target construction, a three-stage causal preprocessing pipeline, two alternative feature engineering strategies, model specifications, a walk-forward backtesting protocol, and evaluation procedures. Figure~\ref{fig:pipeline} provides a schematic overview.\footnote{All transformations are applied in a strictly causal (non-anticipative) manner: every quantity computed at time $t$ uses only information available at or before $t-1$.}

\subsection{Realized Volatility}
\label{sec:rv_definition}

The forecasting target is realized variance (RV) computed at the 30-minute bar level from 5-minute intraday log-returns. For the $j$-th 30-minute bar on trading day $d$, realized volatility is defined as
\begin{equation}
    RV_{d,j} = \sum_{i=1}^{M} r_{d,j,i}^2,
    \label{eq:rv}
\end{equation}
where $r_{d,j,i}$ denotes the $i$-th 5-minute log-return within the bar and $M = 6$ is the number of non-overlapping 5-minute intervals per 30-minute bar. Because data are recorded over 24-hour trading days (including pre-market, regular, and after-hours sessions), each day contains $48$ bars of 30 minutes each. For notational convenience, we index observations sequentially as $RV_t$ where $t$ enumerates consecutive 30-minute bars across all trading days. The forecasting objective is to predict the one-step-ahead realized volatility $RV_{t+1}$.

\subsection{Data Preprocessing}
\label{sec:preprocessing}

Raw realized volatility and exogenous variables undergo a three-stage transformation pipeline applied in the following fixed order:
\begin{equation}
    \text{Raw series} \;\xrightarrow{\text{Stage 1}}\; \text{Diurnal adjustment} \;\xrightarrow{\text{Stage 2}}\; \text{Semantic transform} \;\xrightarrow{\text{Stage 3}}\; \text{Rolling winsorization}.
    \label{eq:pipeline}
\end{equation}
The pipeline is designed to (i) remove predictable intraday patterns, (ii) stabilize distributional properties, and (iii) limit the influence of extreme observations. Importantly, at evaluation time only Stages 1 and 2 are inverted to recover raw-scale forecasts; Stage 3 (winsorization) is a lossy operation that is not reversed.

\subsubsection{Stage 1: Diurnal Adjustment}
\label{sec:diurnal}

Intraday volatility exhibits a well-documented U-shaped pattern, with elevated levels near the market open and close. To remove this deterministic seasonality, we divide each variable by a rolling time-of-day baseline. Let $s(t) \in \{1, \ldots, 48\}$ denote the time-of-day slot for observation $t$. For non-negative variables such as $RV_t$, the diurnal baseline is
\begin{equation}
    \bar{RV}_{t \mid s} = \frac{1}{|\mathcal{W}_t|} \sum_{j \in \mathcal{W}_t} RV_j,
    \label{eq:diurnal_mean}
\end{equation}
where $\mathcal{W}_t = \{j : j < t,\; s(j) = s(t)\}$ denotes the set of up to $W = 20$ most recent observations sharing the same time slot, with a minimum of 5 observations required before the baseline becomes valid. The one-period shift ($j < t$, not $j \leq t$) ensures strict causality. The adjusted series is then
\begin{equation}
    \widetilde{RV}_t = \frac{RV_t}{\bar{RV}_{t \mid s}}.
    \label{eq:diurnal_adj}
\end{equation}
For signed variables (e.g., autocovariance), the baseline is instead the rolling standard deviation within each slot, and the adjustment divides by this rolling standard deviation. Observations in the initial warm-up period (before 5 same-slot observations have accumulated) have their baseline set to 1, effectively passing through unadjusted; these observations are subsequently consumed by the training buffer and do not enter the test set.

\subsubsection{Stage 2: Semantic Transforms}
\label{sec:semantic_transforms}

Following diurnal adjustment, we apply variance-stabilizing transforms chosen according to the distributional properties of each variable class:
\begin{itemize}
    \item Squared-return and volatility measures ($RV$, bipower variation, turnover, effective spread): $x \mapsto \sqrt{x}$.
    \item Autocovariance: $x \mapsto \mathrm{sign}(x)\sqrt{|x|}$.
    \item Cubic returns: $x \mapsto x^{1/3}$.
    \item Quartic returns: $x \mapsto x^{1/4}$.
    \item Implied volatility measures (VIX, VVIX, VIX3M): $x \mapsto \log(x)$.
\end{itemize}
These transforms compress the right tail and reduce heteroskedasticity, which is particularly beneficial for linear models. The square-root transform applied to $RV$ is especially important: since models are trained on the $\sqrt{RV}$ scale, a bias correction (Duan's smearing estimator, Section~\ref{sec:duan}) is required when converting forecasts back to the raw variance scale.

\subsubsection{Stage 3: Rolling Winsorization}
\label{sec:winsorization}

After transformation, each series is clipped to rolling 5th and 95th percentile bounds computed over a trailing window of $W_{\mathrm{clip}} = 240$ observations (approximately 5 trading days):
\begin{equation}
    x_t^{w} = \mathrm{clip}\!\left(x_t;\; q_{0.05}^{(t)},\; q_{0.95}^{(t)}\right),
    \label{eq:winsor}
\end{equation}
where $q_{\alpha}^{(t)}$ denotes the $\alpha$-quantile of the 240-observation trailing window ending at $t-1$. The one-period shift ensures that quantile bounds are computed from past data only. Because winsorization is a lossy compression, it cannot be inverted at evaluation time; forecasts are mapped back to the raw scale using only the inverse of Stages 1--2 (Section~\ref{sec:duan}).

\subsection{Feature Engineering}
\label{sec:features}

We consider two alternative strategies for constructing lagged features from the preprocessed series: a hand-crafted HAR lag structure and a data-driven PCA compression. Both strategies transform each variable's history into a fixed-dimensional feature vector suitable for regression.

\subsubsection{HAR Features}
\label{sec:har_features}

The Heterogeneous Autoregressive (HAR) model of \citet{Corsi2009} captures the multi-scale persistence of volatility through aggregated lagged components. We adopt a geometric base-5 lag structure adapted to the 30-minute sampling frequency, yielding the lag set
\begin{equation}
    \mathcal{L} = \{1, 5, 25, 125, 625, 3125\},
    \label{eq:har_lags}
\end{equation}
which corresponds approximately to horizons of 30 minutes, 2.5 hours, half a trading day, 2.6 trading days, 13 trading days, and 65 trading days. The base-5 geometric spacing provides convenient multi-scale coverage at 30-minute resolution, analogous to the daily/weekly/monthly structure in \citet{Corsi2009} but extended to higher frequencies. For each lag $k \in \mathcal{L}$, the HAR feature is the trailing rolling mean of the preprocessed target:
\begin{equation}
    \overline{RV}_t^{(k)} = \frac{1}{k}\sum_{i=0}^{k-1} \widetilde{RV}_{t-1-i},
    \label{eq:har_ma}
\end{equation}
where the shift by one ensures no contemporaneous information enters the feature. The standard HAR regression model takes the form
\begin{equation}
    \widetilde{RV}_{t+1} = \beta_0 + \sum_{k \in \mathcal{L}} \beta_k \, \overline{RV}_t^{(k)} + \varepsilon_{t+1}.
    \label{eq:har_model}
\end{equation}

When exogenous variables are included, the same HAR lag structure is applied to each additional preprocessed column, generating $|\mathcal{L}| = 6$ rolling-mean features per variable.

\subsubsection{PCA Features}
\label{sec:pca_features}

As an alternative to the fixed HAR lag structure, we consider a data-driven approach that serves two purposes: (i) it lets the data determine the optimal lag weighting rather than imposing the HAR rolling-mean structure, and (ii) it reduces dimensionality when many exogenous variables are included. We first construct approximately 20 log-spaced raw lags spanning the range $[1, 3125]$, providing dense coverage at short horizons and sparser coverage at longer horizons. For each variable, individual shifted lags $x_{t-\ell}$ for $\ell \in \mathcal{L}_{\mathrm{pca}}$ are computed (rather than rolling means as in the HAR approach).

These raw lag features are then projected onto a low-dimensional subspace via Principal Component Analysis (PCA). Specifically, on each rolling training window of length $T_{\mathrm{train}}$, we fit PCA with $n_c = 5$ components using a randomized SVD solver:
\begin{equation}
    \mathbf{z}_t = \mathbf{W}^{\top}(\mathbf{x}_t - \bar{\mathbf{x}}),
    \label{eq:pca}
\end{equation}
where $\mathbf{W} \in \mathbb{R}^{p \times n_c}$ is the projection matrix estimated from the training buffer and $\bar{\mathbf{x}}$ is the training-set mean. The PCA transform is re-estimated each time the model is refit, ensuring the projection adapts to evolving autocorrelation structures.

\subsubsection{Exogenous Variable Subgroups}
\label{sec:subgroups}

To assess the marginal contribution of different information sources, we organize exogenous variables into thematic subgroups. Each subgroup augments the target-only baseline (HAR or PCA features of $\widetilde{RV}$ alone) with additional features derived from the subgroup's variables using the same lag structure. The subgroups are:

\begin{itemize}
    \item \textbf{Baseline}: HAR (or PCA) features of $\widetilde{RV}$ only --- no exogenous variables.
    \item \textbf{Moments}: Higher-order return moments and related measures (bipower variation, realized semi-variance, autocovariance, absolute returns, cubic and quartic returns).
    \item \textbf{Liquidity}: Order flow and market microstructure variables (volume, turnover, buy/sell turnover, effective spread, quoted spread, number of observations).
    \item \textbf{Market (EW)}: Equal-weighted cross-sectional stock factor aggregates (returns, bipower, semi-variance, absolute returns).
    \item \textbf{Market (VW)}: Value-weighted analogues of the equal-weighted factors.
    \item \textbf{Sentiment}: Social media sentiment metrics (StockTwits attention, sentiment score, sentiment count).
    \item \textbf{Implied Volatility}: Options-implied measures (VIX, VVIX, VIX3M).
    \item \textbf{Volatility Demand}: Options order flow measures (SPX open/close, all options open/close, open only).
    \item \textbf{All Features}: The union of all exogenous variables above.
\end{itemize}

\subsection{Models}
\label{sec:models}

We consider five models spanning parametric linear, ensemble tree-based, and naive approaches. All models are trained on the preprocessed (diurnally adjusted, transformed, winsorized) scale. Key differences in refit frequency reflect computational cost: ridge regression admits a closed-form solution and is refit at every time step, while tree ensembles are more expensive and are refit every 5 steps.

\subsubsection{Naive Baseline}
\label{sec:naive}

The naive baseline predicts $\widetilde{RV}_{t+1}$ using the HAR rolling mean at the 125-period horizon (approximately 2.6 trading days):
\begin{equation}
    \hat{y}_{t+1}^{\mathrm{naive}} = \overline{RV}_t^{(125)}.
    \label{eq:naive}
\end{equation}
This provides a non-trivial benchmark that reflects the strong persistence of realized volatility.

\subsubsection{Ridge Regression}
\label{sec:ridge}

Ridge regression augments ordinary least squares with an $\ell_2$ penalty on the coefficient vector:
\begin{equation}
    \hat{\boldsymbol{\beta}} = \arg\min_{\boldsymbol{\beta}} \left\{ \|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|_2^2 + \alpha \|\boldsymbol{\beta}\|_2^2 \right\},
    \label{eq:ridge}
\end{equation}
with regularization parameter $\alpha = 1.0$. Input features are standardized using a rolling robust scaler based on the median and interquartile range (IQR) estimated from the current training buffer. Ridge is refit at every time step via the closed-form solution.

\subsubsection{XGBoost}
\label{sec:xgboost}

We employ XGBoost \citep{Chen2016} as a gradient-boosted tree ensemble with hyperparameters: maximum tree depth of 3, 100 boosting rounds, and a learning rate of 0.1. The histogram-based splitting method is used for computational efficiency. The model is refit every 5 time steps.

\subsubsection{LightGBM}
\label{sec:lightgbm}

LightGBM \citep{Ke2017} is configured with matching hyperparameters to XGBoost (maximum depth 3, 100 estimators, learning rate 0.1) to enable direct comparison. The model is refit every 5 time steps.

\subsubsection{Random Forest}
\label{sec:rf}

A random forest ensemble is trained with 100 trees and a maximum depth of 3. Like the other tree-based models, it is refit every 5 time steps.

\medskip

\noindent\textit{Categorical time features.} All three tree-based models (XGBoost, LightGBM, random forest) additionally receive day-of-week and hour-of-day as categorical feature columns appended to the HAR or PCA features. Ridge regression does not receive these features, as the diurnal adjustment (Stage 1) already removes time-of-day effects from the input series.

\subsection{Walk-Forward Backtesting}
\label{sec:backtest}

All models are evaluated using a strictly out-of-sample walk-forward procedure. At each test period $t$, the model is trained (or updated) on a rolling window of $T_{\mathrm{train}} = 500$ consecutive observations immediately preceding $t$. With 48 bars per 24-hour trading day, this corresponds to approximately 10.4 trading days of training data. The procedure is:

\begin{enumerate}
    \item \textbf{Initialize}: Fit the model on observations $\{t - T_{\mathrm{train}}, \ldots, t-1\}$.
    \item \textbf{Predict}: Generate forecast $\hat{y}_{t}$ using features $\mathbf{x}_t$.
    \item \textbf{Observe}: Record the realized target $y_t$.
    \item \textbf{Update}: Add $(\mathbf{x}_t, y_t)$ to the rolling buffer, drop the oldest observation, and refit the model at the model-specific refit frequency (every step for ridge, every 5 steps for tree models).
    \item \textbf{Advance}: Increment $t$ and return to step 2.
\end{enumerate}

For ridge regression, feature scaling parameters (median and IQR) are updated at every step using the rolling robust scaler fit on the current training buffer. For PCA-based configurations, the projection matrix $\mathbf{W}$ is re-estimated at each refit. This ensures all model components adapt to non-stationarities while maintaining strict temporal separation between training and test data.

The backtest is parallelized across 100 chunks via SLURM array jobs, with each chunk processing a contiguous subset of test indices.

\subsection{Evaluation}
\label{sec:evaluation}

\subsubsection{Inverse Transform and Duan's Smearing Estimator}
\label{sec:duan}

Models are trained on the preprocessed scale (after diurnal adjustment, semantic transform, and winsorization). To evaluate forecasts against raw realized volatility, predictions must be mapped back through the inverse of the preprocessing pipeline. Of the three stages, only Stages 1 (diurnal adjustment) and 2 (semantic transform) are invertible; Stage 3 (winsorization) is lossy and is not reversed.

For the primary target $RV$, the semantic transform is the square root ($\sqrt{\cdot}$), so the na\"ive inverse would simply square the forecast: $\hat{y}_t^2$. However, because $\mathbb{E}[\sqrt{X}]^2 \neq \mathbb{E}[X]$ by Jensen's inequality, squaring a forecast of $\sqrt{RV}$ produces a downward-biased estimate of $RV$. Duan's smearing estimator \citep{Duan1983} corrects this bias by adding the in-sample residual variance before inverting:
\begin{equation}
    \widehat{RV}_t^{\mathrm{raw}} = \left(\hat{y}_t^2 + \hat{\sigma}_{\varepsilon}^2\right) \cdot \bar{RV}_{t \mid s}^{\mathrm{baseline}},
    \label{eq:duan}
\end{equation}
where $\hat{\sigma}_{\varepsilon}^2 = \frac{1}{N}\sum_{i}(y_i - \hat{y}_i)^2$ is the in-sample mean squared residual (estimating the variance lost by working on the transformed scale) and $\bar{RV}_{t \mid s}^{\mathrm{baseline}}$ is the diurnal baseline factor at time $t$ that reverses Stage 1.

\subsubsection{QLIKE Loss}
\label{sec:qlike}

Our primary evaluation metric is the quasi-likelihood (QLIKE) loss function, which is the robust loss function for volatility forecast comparison under general loss functions \citep{Patton2011}:
\begin{equation}
    L_{\mathrm{QLIKE}} = \frac{1}{N}\sum_{t=1}^{N} \left[ \frac{RV_t^{\mathrm{raw}}}{\widehat{RV}_t^{\mathrm{raw}}} - \log\!\left(\frac{RV_t^{\mathrm{raw}}}{\widehat{RV}_t^{\mathrm{raw}}}\right) - 1 \right].
    \label{eq:qlike}
\end{equation}
QLIKE is strictly consistent for the conditional mean under the assumption that realized volatility is an unbiased estimator of integrated variance \citep{Patton2011}. It penalizes under-prediction more heavily than over-prediction, reflecting the asymmetric costs relevant in risk management applications.

\subsubsection{Additional Metrics}
\label{sec:additional_metrics}

We also report mean squared error (MSE) and mean absolute error (MAE) computed on the preprocessed (adjusted) scale. To limit the influence of extreme outliers, winsorized variants (clipping errors to the 5th--95th percentile range) are computed for all metrics. Out-of-sample $R^2$ is defined relative to the naive baseline:
\begin{equation}
    R^2_{\mathrm{OOS}} = 1 - \frac{\mathrm{MSE}_{\mathrm{model}}}{\mathrm{MSE}_{\mathrm{naive}}},
    \label{eq:oos_r2}
\end{equation}
where positive values indicate improvement over the naive forecast.

\subsection{Experimental Design}
\label{sec:experimental_design}

We report results from three complementary analyses, each designed to isolate a specific dimension of the forecasting problem:

\begin{enumerate}
    \item \textbf{Ridge--HAR Subgroup Analysis.} Ridge regression with HAR features is estimated for each exogenous subgroup (Section~\ref{sec:subgroups}). By holding the model fixed (ridge) and varying the information set, this analysis isolates the marginal predictive content of each information source beyond the target's own history.

    \item \textbf{HAR Model Comparison.} Using only the target HAR features (baseline subgroup, no exogenous variables), we compare all five models --- naive, ridge, XGBoost, LightGBM, and random forest. By holding the features fixed and varying the model, this analysis assesses whether nonlinear models capture dynamics beyond the linear HAR specification.

    \item \textbf{Ridge--PCA Subgroup Analysis.} Ridge regression with PCA features is estimated for each subgroup, paralleling the HAR analysis. By replacing the hand-crafted HAR lag structure with data-driven PCA compression while keeping the model fixed, this analysis evaluates whether flexible lag weighting and dimensionality reduction offer complementary predictive power.
\end{enumerate}
